### 2022-12-12 lmcs/conclusion.tex
removing some emphs

While our work focuses on \emph{binary} session types, it is worth discussing related work on Multiparty Session Types (MPST), which describe communication among multiple agents in a distributed setting. The line of work on MPST starts with Honda~\etal~\cite{hondayoshida08}, which guarantees deadlock freedom \emph{within a single session} by design, but the property does not hold for session interleaving.
Bettini~\etal~\cite{bettinicoppo08} follow a technique similar to Kobayashi's for MPST.
The main difference with our work is that we associate priorities with communication actions, where Bettini~\etal~\cite{bettinicoppo08} associate them with channels.
Coppo \etal \cite{coppoetal13,coppoetal16} present a deterministic, sound and complete, and compositional type inference algorithm for an \emph{interaction type system}, which is used to guarantee global progress for processes in a calculus based on asynchronous as well as dynamically interleaved and interfered multiparty sessions. The interaction type system allows to infer \emph{causalities}---much in the line of Kobayashi's priorities ---of channels, thus guaranteeing that session-typed processes do not get stuck at intermediate stages of their sessions.
Carbone and Montesi~\cite{carbonemontesi13} combine MPST with choreographic programming and obtain a formalism that satisfies deadlock freedom. In the same vein as MPST, choreographic programming specifies communication among all participants in a distributed system. While MPST target mainly protocol descriptions, choreographies have mainly targetted implementations and programming languages as they are suitable for describing concrete system implementations.
Deni\'{e}lou and Yoshida~\cite{DenielouY13} introduce \emph{multiparty compatibility}, which generalises the notion of duality in binary session types. They synthesise safe and deadlock-free \emph{global types}--specifying communication among \emph{all} involved participants, from \emph{local session types}--specifying communication from the viewpoint of \emph{one} participant. To do so, they leverage Labelled Transition Systems (LTSs) and communicating automata.
Castellani~\etal~\cite{CastellaniDGH20} guarantee livelock freedom, a stronger property than deadlock freedom, for MPST with \emph{internal delegation}, where participants in the same session are allowed to delegate tasks to each other, and internal delegation is captured by the global type.
Scalas and Yoshida~\cite{scalasyoshida19} provide a revision of the foundations for MPST, and offer a less complicated and more general theory, by removing duality/consistency. The type systems is parametric and type checking is decidable, but allows for a novel integration of model checking techniques. More protocols and processes can be typed and are guaranteed to be free of deadlocks.

@@

\subsubsection*{Ties with logic}
The correspondence between logic and types lays the foundation for functional programming~\cite{wadler15}.
Since its inception by Girard~\cite{girard87}, linear logic has been a candidate for a foundational correspondence for concurrent programs.
A~correspondence with linear $\pi$-calculus was established early on by Abramsky~\cite{abramsky94} and Bellin and Scott~\cite{bellinscott94}. Many years later, several correspondences between linear logic and the $\pi$-calculus with binary session types were proposed. Caires and Pfenning~\cite{cairespfenning10} propose a correspondence with dual intuitionistic linear logic, while Wadler~\cite{wadler12} proposes a correspondence with classical linear logic. Both works guarantee deadlock freedom as a consequence of adopting cut elimination.
Building on a previous work \cite{cairespfenning10}, Toninho \etal \cite{toninhoetal13} present a Curry-Howard correspondence between session types and linear logic for functional language via linear contextual monads, which are first-class values, thus giving rise to a higher-order session-typed  language. In addition to the more standard results, the authors also prove a global progress theorem.
Dardha and Gay~\cite{dardhagay18extended} define Priority CP by integrating Kobayashi and Padovani's work on priorities~\cite{kobayashi06,padovani14} with CP, loosening its ties to linear logic in exchange for expressivity. However, they show how PCP can be also \emph{viewed as} a an extension of linear logic, which they call Priority Linear Logic (PLL), and uses mix and cycle rules as opposed to the cut rule.
Dardha and P\'{e}rez~\cite{dardhaperez15extended,dardhaperez15,DardhaP22} compare priorities \`a la Kobayashi with tree restrictions \`a la CP, and show that the latter is a subsystem of the former. In addition, they give a detailed account of comparing several type systems for deadlock freedom spanning across session types, linear logic, and linear types.
Carbone~\etal~\cite{CarboneMSY15,carbonelindley16} give a logical view of MPST with a generalised duality.
Caires and P\'{e}rez~\cite{CairesP16} give a presentation of MPST in terms of binary session types and the use of a \emph{medium process} which guarantee protocol fidelity and deadlock freedom. Their binary session types are rooted in linear logic.
Ciobanu and Horne~\cite{CiobanuH15} give the first Curry-Howard correspondence between MPST and BV~\cite{Guglielmi07}, a conservative extension of linear logic with a non-commutative operator for sequencing.
Horne~\cite{Horne20} give a system for subtyping and multiparty compatibility where compatible processes are race free and deadlock free using a Curry-Howard correspondence, similar to the approach in~\cite{CiobanuH15}.
Balzer~\etal~\cite{balzerpfenning17} introduce sharing at the cost of deadlock freedom, which they restore using \emph{worlds}, an approach similar to priorities~\cite{balzertoninho19}.
Staying on sharing, Rocha and Caires \cite{rochacaires21} introduce an imperative feature, that of shared mutable states into a functional language based on Curry-Howard correspondence with linear logic. Their type system is thus able to capture programs which were not possible in previous works. The authors prove extensive technical results, including session fidelity, progress, confluence and
normalisation.
Lastly, Jacobs \etal \cite{JacobsBK22a} present a novel technique to guarantee deadlock freedom based on the notion of connectivity graph, which is an abstract representation of the topology of concurrent systems, and separation logic used to substructurally treat connectivity graph edges and labels.

>>>>

While our work focuses on {binary} session types, it is worth discussing related work on Multiparty Session Types (MPST), which describe communication among multiple agents in a distributed setting. The line of work on MPST starts with Honda~\etal~\cite{hondayoshida08}, which guarantees deadlock freedom {within a single session} by design, but the property does not hold for session interleaving.
Bettini~\etal~\cite{bettinicoppo08} follow a technique similar to Kobayashi's for MPST.
The main difference with our work is that we associate priorities with communication actions, where Bettini~\etal~\cite{bettinicoppo08} associate them with channels.
Coppo \etal \cite{coppoetal13,coppoetal16} present a deterministic, sound and complete, and compositional type inference algorithm for an \emph{interaction type system}, which is used to guarantee global progress for processes in a calculus based on asynchronous as well as dynamically interleaved and interfered multiparty sessions. The interaction type system allows to infer \emph{causalities}---much in the line of Kobayashi's priorities ---of channels, thus guaranteeing that session-typed processes do not get stuck at intermediate stages of their sessions.
Carbone and Montesi~\cite{carbonemontesi13} combine MPST with choreographic programming and obtain a formalism that satisfies deadlock freedom. In the same vein as MPST, choreographic programming specifies communication among all participants in a distributed system. While MPST target mainly protocol descriptions, choreographies have mainly targetted implementations and programming languages as they are suitable for describing concrete system implementations.
Deni\'{e}lou and Yoshida~\cite{DenielouY13} introduce \emph{multiparty compatibility}, which generalises the notion of duality in binary session types. They synthesise safe and deadlock-free \emph{global types}--specifying communication among {all} involved participants, from \emph{local session types}--specifying communication from the viewpoint of {one} participant. To do so, they leverage Labelled Transition Systems (LTSs) and communicating automata.
Castellani~\etal~\cite{CastellaniDGH20} guarantee livelock freedom, a stronger property than deadlock freedom, for MPST with {internal delegation}, where participants in the same session are allowed to delegate tasks to each other, and internal delegation is captured by the global type.
Scalas and Yoshida~\cite{scalasyoshida19} provide a revision of the foundations for MPST, and offer a less complicated and more general theory, by removing duality/consistency. The type systems is parametric and type checking is decidable, but allows for a novel integration of model checking techniques. More protocols and processes can be typed and are guaranteed to be free of deadlocks.

@@

\subsubsection*{Ties with logic}
The correspondence between logic and types lays the foundation for functional programming~\cite{wadler15}.
Since its inception by Girard~\cite{girard87}, linear logic has been a candidate for a foundational correspondence for concurrent programs.
A~correspondence with linear $\pi$-calculus was established early on by Abramsky~\cite{abramsky94} and Bellin and Scott~\cite{bellinscott94}. Many years later, several correspondences between linear logic and the $\pi$-calculus with binary session types were proposed. Caires and Pfenning~\cite{cairespfenning10} propose a correspondence with dual intuitionistic linear logic, while Wadler~\cite{wadler12} proposes a correspondence with classical linear logic. Both works guarantee deadlock freedom as a consequence of adopting cut elimination.
Building on a previous work \cite{cairespfenning10}, Toninho \etal \cite{toninhoetal13} present a Curry-Howard correspondence between session types and linear logic for functional language via linear contextual monads, which are first-class values, thus giving rise to a higher-order session-typed  language. In addition to the more standard results, the authors also prove a global progress theorem.
Dardha and Gay~\cite{dardhagay18extended} define Priority CP by integrating Kobayashi and Padovani's work on priorities~\cite{kobayashi06,padovani14} with CP, loosening its ties to linear logic in exchange for expressivity. However, they show how PCP can be also {viewed as} a an extension of linear logic, which they call Priority Linear Logic (PLL), and uses mix and cycle rules as opposed to the cut rule.
Dardha and P\'{e}rez~\cite{dardhaperez15extended,dardhaperez15,DardhaP22} compare priorities \`a la Kobayashi with tree restrictions \`a la CP, and show that the latter is a subsystem of the former. In addition, they give a detailed account of comparing several type systems for deadlock freedom spanning across session types, linear logic, and linear types.
Carbone~\etal~\cite{CarboneMSY15,carbonelindley16} give a logical view of MPST with a generalised duality.
Caires and P\'{e}rez~\cite{CairesP16} give a presentation of MPST in terms of binary session types and the use of a \emph{medium process} which guarantee protocol fidelity and deadlock freedom. Their binary session types are rooted in linear logic.
Ciobanu and Horne~\cite{CiobanuH15} give the first Curry-Howard correspondence between MPST and BV~\cite{Guglielmi07}, a conservative extension of linear logic with a non-commutative operator for sequencing.
Horne~\cite{Horne20} give a system for subtyping and multiparty compatibility where compatible processes are race free and deadlock free using a Curry-Howard correspondence, similar to the approach in~\cite{CiobanuH15}.
Balzer~\etal~\cite{balzerpfenning17} introduce sharing at the cost of deadlock freedom, which they restore using \emph{worlds}, an approach similar to priorities~\cite{balzertoninho19}.
Staying on sharing, Rocha and Caires \cite{rochacaires21} introduce an imperative feature, that of shared mutable states into a functional language based on Curry-Howard correspondence with linear logic. Their type system is thus able to capture programs which were not possible in previous works. The authors prove extensive technical results, including session fidelity, progress, confluence and
normalisation.
Lastly, Jacobs \etal \cite{JacobsBK22a} present a novel technique to guarantee deadlock freedom based on the notion of connectivity graph, which is an abstract representation of the topology of concurrent systems, and separation logic used to substructurally treat connectivity graph edges and labels.

### 2022-12-12 lmcs/introduction.tex
removing some emphs

GV is a concurrent $\lambda$-calculus with \emph{binary} session types, where each channel is shared between exactly two processes. Binary session types guarantee two crucial properties: \emph{communication safety}---\eg, if the protocol says to transmit an integer, you transmit an integer---and \emph{session fidelity}---\eg, if the protocol says send, you send. A third crucial property is \emph{deadlock freedom}, which ensures that processes do not have cyclic dependencies---\eg, when two processes wait for each other to send a value. Binary session types \emph{alone} are insufficient to rule out deadlocks arising from interleaved sessions, but several additional techniques have been developed to guarantee deadlock freedom in session-typed $\pi$-calculus and concurrent $\lambda$-calculus.

>>>>

GV is a concurrent $\lambda$-calculus with \emph{binary} session types, where each channel is shared between exactly two processes. Binary session types guarantee two crucial properties: \emph{communication safety}---\eg, if the protocol says to transmit an integer, you transmit an integer---and \emph{session fidelity}---\eg, if the protocol says send, you send. A third crucial property is \emph{deadlock freedom}, which ensures that processes do not have cyclic dependencies---\eg, when two processes wait for each other to send a value. Binary session types {alone} are insufficient to rule out deadlocks arising from interleaved sessions, but several additional techniques have been developed to guarantee deadlock freedom in session-typed $\pi$-calculus and concurrent $\lambda$-calculus.

### 2022-12-12 lmcs/priority-cp-long.tex
removing some emphs

Types $\pcp{\ty{\tynil[\cs{o}]}}$ and $\pcp{\ty{\tytop[\cs{o}]}}$ type the endpoints of a channel over which we can send or receive a choice between \emph{no options}. These two types act as units for $\pcp{\ty{\typlus[\cs{o}]{A}{B}}}$ and $\pcp{\ty{\tywith[\cs{o}]{A}{B}}}$, respectively.

@@

The main change we make to PCP is \emph{removing commuting conversions}. Commuting conversions are necessary if we want our reduction strategy to correspond \emph{exactly} to cut (or cycle in \cite{dardhagay18extended}) elimination. However, as Lindley and Morris~\cite{lindleymorris15} show, all communications that can be performed \emph{with} the use of commuting conversions, can also be performed \emph{without} them, but using structural congruence.

>>>>

Types $\pcp{\ty{\tynil[\cs{o}]}}$ and $\pcp{\ty{\tytop[\cs{o}]}}$ type the endpoints of a channel over which we can send or receive a choice between {no options}. These two types act as units for $\pcp{\ty{\typlus[\cs{o}]{A}{B}}}$ and $\pcp{\ty{\tywith[\cs{o}]{A}{B}}}$, respectively.

@@

The main change we make to PCP is {removing commuting conversions}. Commuting conversions are necessary if we want our reduction strategy to correspond {exactly} to cut (or cycle in \cite{dardhagay18extended}) elimination. However, as Lindley and Morris~\cite{lindleymorris15} show, all communications that can be performed \emph{with} the use of commuting conversions, can also be performed \emph{without} them, but using structural congruence.

### 2022-12-12 lmcs/priority-gv.tex
polishing text
removing some emphs

We present Priority GV (PGV), a session-typed functional language based on GV~\cite{wadler15,lindleymorris15} which uses priorities \`{a} la Kobayashi and Padovani~\cite{kobayashi06,padovaninovara15} to enforce deadlock freedom.
Priority GV offers a more fine-grained analysis of communication structures, and by separating channel creation form thread spawning it allows cyclic structures.

@@

Types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
Type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, $\cs{p}$ is a \emph{lower bound} on the priorities of the endpoints captured by the body of the function, and $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function. The type of \emph{pure functions} $\ty{\tylolli{T}{U}}$, \ie those which perform no communications, is syntactic sugar for $\ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}$. The lower bound for a pure function is $\cs{\ptop}$ as pure functions never start communicating. For similar reasons, the upper bound for a pure function is $\cs{\pbot}$.

@@

\subsubsection*{Type Duality}
Duality plays a crucial role in session types. The two endpoints of a channel are assigned dual types, ensuring that, for instance, whenever one program \emph{sends} a value on a channel, the program on the other end is waiting to \emph{receive}. Each session type $\ty{S}$ has a dual, written $\ty{\co{S}}$. Duality is an involutive function which \emph{preserves priorities}:
\[
  \ty{\co{\tysend[\cs{o}]{T}{S}}} = \ty{\tyrecv[\cs{o}]{T}{\co{S}}}
  \qquad
  \ty{\co{\tyrecv[\cs{o}]{T}{S}}} = \ty{\tysend[\cs{o}]{T}{\co{S}}}
  \qquad
  \ty{\co{\tyends[\cs{o}]}} = \ty{\tyendr[\cs{o}]}
  \qquad
  \ty{\co{\tyendr[\cs{o}]}} = \ty{\tyends[\cs{o}]}
\]

@@

Term reduction is the standard call-by-value, left-to-right evaluation for GV, and only deviates from reduction for the linear $\lambda$-calculus in that it reduces terms to values \emph{or} ready terms waiting to perform a communication action.

@@

Intuitively, evaluating $\tm{\link}$ causes a substitution: if we have a channel bound as $\tm{\res{x}{y}{}}$, then $\tm{\link\;\pair{w}{x}}$ replaces all occurrences of $\tm{y}$ by $\tm{w}$. However, in GV, $\tm{\link}$ is required to return a terminated channel, which means that the semantics for $\tm{link}$ must \emph{create} a fresh channel of type $\ty{\tyends}/\ty{\tyendr}$. The endpoint of type $\ty{\tyends}$ is returned by the $\tm{link}$ construct, and a $\tm{\wait}$ on the other endpoint guards the \emph{actual} substitution. In PGV, evaluating $\tm{\link}$ simply causes a substitution.
\item
\emph{Our structural congruence is type preserving}. Consequently, we can embed it directly into the reduction relation. In GV, this is not the case, and subject reduction relies on proving that if $\equiv\cred$ ends up in an ill-typed configuration, we can rewrite it to a well typed configuration using $\equiv$.
\end{enumerate}

@@

\subsubsection*{Terms Typing}
%\Cref{fig:pgv-typing} gives the typing rules for PGV.
Typing rules for terms are at the top of \cref{fig:pgv-typing}.
Terms are typed by a judgement $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ stating that ``a term $\tm{M}$ has type $\ty{T}$ and an upper bound on its priority $\cs{p}$ under the typing environment $\ty{\Gamma}$''. Typing for the linear $\lambda$-calculus is standard. Linearity is ensured by splitting environments on branching rules, requiring that the environment in the variable rule consists of just the variable, and the environment in the constant and unit rules are empty. Constants $\tm{K}$ are typed using type schemas, and embedded using \LabTirName{T-Const} (mid of \cref{fig:pgv-typing}). The typing rules treat \emph{all variables} as linear resources, even those of non-linear types such as $\ty{\tyunit}$, though they can easily be extended to allow values with unrestricted usage~\cite{wadler14}.

The only non-standard feature of the typing rules is the priority annotations. Priorities are based on \emph{obligations/capabilities} used by Kobayashi~\cite{kobayashi06}, and simplified to single priorities following Padovani~\cite{padovani14}. The integration of priorities into GV is adapted from Padovani and Novara~\cite{padovaninovara15}. Paraphrasing Dardha and Gay~\cite{dardhagay18extended}, priorities obey the following two laws:
\begin{enumerate}[label= (\roman*) ]
\item an action with lower priority happens before an action with higher priority; and
\item communication requires \emph{equal} priorities for dual actions.
\end{enumerate}

In PGV, we keep track of a lower and upper bound on the priorities of a term, \ie, while evaluating the term, when it starts communicating, and when it finishes, respectively. The upper bound is written on the sequent and the lower bound is approximated from the typing environment, \eg, for $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ the upper bound is $\cs{p}$ and the lower bound is at least $\minpr(\ty{\Gamma})$. The latter is \emph{correct} because a term cannot communicate at a priority earlier than the earliest priority amongst the channels it has access to. It is an \emph{approximation} on function terms, as these can ``skip'' communication by returning the corresponding channel unused. However, linearity prevents such functions from being used in well-typed configurations: once the unused channel's priority has passed, it can no longer be used.

@@

\subsubsection*{Configurations Typing}
Typing rules for configurations are at the bottom of \cref{fig:pgv-typing}. Configurations are typed by a judgement $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ stating that ``a configuration $\tm{\conf{C}}$ with flag $\phi$ is well typed under typing environment $\ty{\Gamma}$''. Configuration typing is based on the standard typing for GV. Terms are embedded either as main or as child threads. The priority bound from the term typing is discarded, as configurations contain no further blocking actions. Main threads are allowed to return a value, whereas child threads are required to return the unit value. Sequents are annotated with a flag $\phi$, which ensures that there is at most one main thread.

@@

\subsection{Progress and Deadlock Freedom}
PGV satisfies progress, as PGV configurations either reduce or are in normal form. However, the normal forms may seem surprising at first, as evaluating a well typed PGV term does not necessarily produce \emph{just} a value. If a term returns an endpoint, then its normal form contains a thread which is ready to communicate on the dual of that endpoint. This behaviour is not new to PGV.

@@

If we reduce the above program, we get $\tm{\res{x}{x'}{(\ppar{\child\;\echo_{x}}{\main\;x'})}}$. Clearly, no more evaluation is possible, even though the configuration contains the thread $\tm{\child\;\echo_{x}}$, which is blocked on $\tm{x}$. In~\cref{cor:pgv-closed-progress} we will show that if a term does not return an endpoint, it must produce \emph{only} a value.

@@

\emph{Ready terms} are terms which perform communication actions, either by themselves, \eg creating a new channel or thread, or with another thread, \eg sending or receiving.
It is worth mentioning that the notion of readiness presented here is akin to \emph{live} processes  introduced by Caires and Pfenning \cite{cairespfenning10,DardhaP22}, and \emph{poised} processes introduced by Pfenning and Griffith \cite{PfenningG15} and later used by Balzer \etal \cite{balzerpfenning17,balzertoninho19}. Ready processes like live/poised processes denote processes that are ready to communicate on their providing channel.
\begin{definition}[Ready Terms]%
  \label{def:pgv-ready-actions}
  A~term $\tm{L}$ is \emph{ready} if it is of the form $\tm{\plug{E}{M}}$, where $\tm{M}$ is of the form $\tm{\new}$, $\tm{\spawn\;N}$, $\tm{\link\;\pair{x}{y}}$, or $\tm{M}$ acts on $\tm{x}$. In the latter case, we say that $\tm{L}$ is \emph{ready to act on} $\tm{x}$.
\end{definition}

@@

A~configuration either reduces, or it is equivalent to configuration in normal form. Crucial to the normal form is that each term $\tm{M_i}$ is blocked on the corresponding channel $\tm{x_i}$, and hence no two terms act on dual endpoints. Furthermore, no term $\tm{M_i}$ can perform a communication action by itself, since those are excluded by the definition of actions.
Finally, as a corollary, we get that well typed terms which do not return endpoints return \emph{just} a value:

>>>>

We present Priority GV (PGV), a session-typed functional language based on GV~\cite{wadler15,lindleymorris15} which uses priorities \`{a} la Kobayashi and Padovani~\cite{kobayashi06,padovaninovara15} to enforce deadlock freedom.
Priority GV is more flexible than GV because it allows processes to share more than one communication channel.

@@

Types $\ty{\typrod{T}{U}}$, $\ty{\tyunit}$, $\ty{\tysum{T}{U}}$, and $\ty{\tyvoid}$ are the standard linear $\lambda$-calculus product type, unit type, sum type, and empty type.
Type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$ is the standard linear function type, annotated with \emph{priority bounds} $\cs{p},\cs{q}\in\mathbb{N}\cup\{\cs{\pbot},\cs{\ptop}\}$.
Every session type is also a type.
Given a function with type $\ty{\tylolli[\cs{p},\cs{q}]{T}{U}}$, $\cs{p}$ is a \emph{lower bound} on the priorities of the endpoints captured by the body of the function, and $\cs{q}$ is an \emph{upper bound} on the priority of the communications that take place as a result of applying the function. The type of {pure functions} $\ty{\tylolli{T}{U}}$, \ie those which perform no communications, is syntactic sugar for $\ty{\tylolli[\cs{\ptop},\cs{\pbot}]{T}{U}}$. The lower bound for a pure function is $\cs{\ptop}$ as pure functions never start communicating. For similar reasons, the upper bound for a pure function is $\cs{\pbot}$.

@@

\subsubsection*{Type Duality}
Duality plays a crucial role in session types. The two endpoints of a channel are assigned dual types, ensuring that, for instance, whenever one program {sends} a value on a channel, the program on the other end is waiting to {receive}. Each session type $\ty{S}$ has a dual, written $\ty{\co{S}}$. Duality is an involutive function which {preserves priorities}:
\[
  \ty{\co{\tysend[\cs{o}]{T}{S}}} = \ty{\tyrecv[\cs{o}]{T}{\co{S}}}
  \qquad
  \ty{\co{\tyrecv[\cs{o}]{T}{S}}} = \ty{\tysend[\cs{o}]{T}{\co{S}}}
  \qquad
  \ty{\co{\tyends[\cs{o}]}} = \ty{\tyendr[\cs{o}]}
  \qquad
  \ty{\co{\tyendr[\cs{o}]}} = \ty{\tyends[\cs{o}]}
\]

@@

Term reduction is the standard call-by-value, left-to-right evaluation for GV, and only deviates from reduction for the linear $\lambda$-calculus in that it reduces terms to values {or} ready terms waiting to perform a communication action.

@@

Intuitively, evaluating $\tm{\link}$ causes a substitution: if we have a channel bound as $\tm{\res{x}{y}{}}$, then $\tm{\link\;\pair{w}{x}}$ replaces all occurrences of $\tm{y}$ by $\tm{w}$. However, in GV, $\tm{\link}$ is required to return a terminated channel, which means that the semantics for $\tm{link}$ must {create} a fresh channel of type $\ty{\tyends}/\ty{\tyendr}$. The endpoint of type $\ty{\tyends}$ is returned by the $\tm{link}$ construct, and a $\tm{\wait}$ on the other endpoint guards the {actual} substitution. In PGV, evaluating $\tm{\link}$ simply causes a substitution.
\item
\emph{Our structural congruence is type preserving}. Consequently, we can embed it directly into the reduction relation. In GV, this is not the case, and subject reduction relies on proving that if $\equiv\cred$ ends up in an ill-typed configuration, we can rewrite it to a well typed configuration using $\equiv$.
\end{enumerate}

@@

\subsubsection*{Terms Typing}
%\Cref{fig:pgv-typing} gives the typing rules for PGV.
Typing rules for terms are at the top of \cref{fig:pgv-typing}.
Terms are typed by a judgement $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ stating that ``a term $\tm{M}$ has type $\ty{T}$ and an upper bound on its priority $\cs{p}$ under the typing environment $\ty{\Gamma}$''. Typing for the linear $\lambda$-calculus is standard. Linearity is ensured by splitting environments on branching rules, requiring that the environment in the variable rule consists of just the variable, and the environment in the constant and unit rules are empty. Constants $\tm{K}$ are typed using type schemas, and embedded using \LabTirName{T-Const} (mid of \cref{fig:pgv-typing}). The typing rules treat {all variables} as linear resources, even those of non-linear types such as $\ty{\tyunit}$, though they can easily be extended to allow values with unrestricted usage~\cite{wadler14}.

The only non-standard feature of the typing rules is the priority annotations. Priorities are based on {obligations/capabilities} used by Kobayashi~\cite{kobayashi06}, and simplified to single priorities following Padovani~\cite{padovani14}. The integration of priorities into GV is adapted from Padovani and Novara~\cite{padovaninovara15}. Paraphrasing Dardha and Gay~\cite{dardhagay18extended}, priorities obey the following two laws:
\begin{enumerate}[label= (\roman*) ]
\item an action with lower priority happens before an action with higher priority; and
\item communication requires \emph{equal} priorities for dual actions.
\end{enumerate}

In PGV, we keep track of a lower and upper bound on the priorities of a term, \ie, while evaluating the term, when it starts communicating, and when it finishes, respectively. The upper bound is written on the sequent and the lower bound is approximated from the typing environment, \eg, for $\tseq[\cs{p}]{\ty{\Gamma}}{M}{T}$ the upper bound is $\cs{p}$ and the lower bound is at least $\minpr(\ty{\Gamma})$. The latter is {correct} because a term cannot communicate at a priority earlier than the earliest priority amongst the channels it has access to. It is an \emph{approximation} on function terms, as these can ``skip'' communication by returning the corresponding channel unused. However, linearity prevents such functions from being used in well-typed configurations: once the unused channel's priority has passed, it can no longer be used.

@@

\subsubsection*{Configurations Typing}
Typing rules for configurations are at the bottom of \cref{fig:pgv-typing}. Configurations are typed by a judgement $\cseq[\phi]{\ty{\Gamma}}{\conf{C}}$ stating that ``a configuration $\tm{\conf{C}}$ with flag $\tm\phi$ is well typed under typing environment $\ty{\Gamma}$''. Configuration typing is based on the standard typing for GV. Terms are embedded either as main or as child threads. The priority bound from the term typing is discarded, as configurations contain no further blocking actions. Main threads are allowed to return a value, whereas child threads are required to return the unit value. Sequents are annotated with a flag $\tm\phi$, which ensures that there is at most one main thread.

@@

\subsection{Progress and Deadlock Freedom}
PGV satisfies progress, as PGV configurations either reduce or are in normal form. However, the normal forms may seem surprising at first, as evaluating a well typed PGV term does not necessarily produce {just} a value. If a term returns an endpoint, then its normal form contains a thread which is ready to communicate on the dual of that endpoint. This behaviour is not new to PGV.

@@

If we reduce the above program, we get $\tm{\res{x}{x'}{(\ppar{\child\;\echo_{x}}{\main\;x'})}}$. Clearly, no more evaluation is possible, even though the configuration contains the thread $\tm{\child\;\echo_{x}}$, which is blocked on $\tm{x}$. In~\cref{cor:pgv-closed-progress} we will show that if a term does not return an endpoint, it must produce {only} a value.

@@

\emph{Ready terms} are terms which perform communication actions, either by themselves, \eg creating a new channel or thread, or with another thread, \eg sending or receiving.
It is worth mentioning that the notion of readiness presented here is akin to {live} processes introduced by Caires and Pfenning \cite{cairespfenning10,DardhaP22}, and {poised} processes introduced by Pfenning and Griffith \cite{PfenningG15} and later used by Balzer \etal \cite{balzerpfenning17,balzertoninho19}. Ready processes like live/poised processes denote processes that are ready to communicate on their providing channel.
\begin{definition}[Ready Terms]%
  \label{def:pgv-ready-actions}
  A~term $\tm{L}$ is \emph{ready} if it is of the form $\tm{\plug{E}{M}}$, where $\tm{M}$ is of the form $\tm{\new}$, $\tm{\spawn\;N}$, $\tm{\link\;\pair{x}{y}}$, or $\tm{M}$ acts on $\tm{x}$. In the latter case, we say that $\tm{L}$ is \emph{ready to act on} $\tm{x}$.
\end{definition}

@@

A~configuration either reduces, or it is equivalent to configuration in normal form. Crucial to the normal form is that each term $\tm{M_i}$ is blocked on the corresponding channel $\tm{x_i}$, and hence no two terms act on dual endpoints. Furthermore, no term $\tm{M_i}$ can perform a communication action by itself, since those are excluded by the definition of actions.
Finally, as a corollary, we get that well typed terms which do not return endpoints return {just} a value:
